Magnetised Polish doughnuts revisited

### 2017-06-29 paper_revised/paper_revised.tex
work on the paper

Next, we choose all the initial radial values for the integration of Eq.~\eqref{eq:F} to lie between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$ ($\theta = \pi / 2$). Since we are only interested in the equipotential surfaces inside the Roche lobe of the disc, our choice of initial values provides us a mapping of the equipotential surfaces of the torus. Given that we have already obtained both the equipotential surfaces $\theta(r)$ which cross the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ at the equatorial plane and the values of the potential in that segment, we can obtain the complete potential distribution for the torus (outside of that segment). Once we have the potential distribution, we can find the {\bf fluid} pressure at the centre of the disc from Eq.~\eqref{eq:enthalpy_eq},
\begin{equation}
p_{\mathrm{c}} = w_{\mathrm{c}}(W_{\mathrm{in}} - W_{\mathrm{c}})\left(\frac{\kappa}{\kappa - 1} + \frac{\lambda
}{\lambda
 - 1}\frac{1}{\beta_{\mathrm{m}_{\mathrm{c}}}}\right)^{-1},
\end{equation}
where $w_{\mathrm{c}}$ is the enthalpy at the centre and
\begin{equation}
\label{eq:beta_eq}
\beta_{\mathrm{m}} = p/p_{\mathrm{m}},
\end{equation}
is the magnetisation parameter ($\beta_{\mathrm{m}_{\mathrm{c}}}$ being the magnetisation parameter at the centre of the disc). Using this definition, we can find the magnetic pressure at the centre,
\begin{equation}
p_{\mathrm{m_{\mathrm{c}}}} = p_{\mathrm{c}}/\beta_{\mathrm{m}_{\mathrm{c}}}\,.
\end{equation}
With both pressures known at the centre, fluid and magnetic, we can now find the constants $K$ and $K_{\mathrm{m}}$ using equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag}. Therefore, for a given inner radius of the disc $r_{\mathrm{in}}$ we can obtain the potential $W_{\mathrm{in}}$. We now finally have all the ingredients required to find the enthalpy distribution (Eq.~\eqref{eq:enthalpy_eq}), the fluid pressure and magnetic pressure distributions (Eqs.~\eqref{eq:eos_fluid} and \eqref{eq:eos_mag}), and the {\bf rest-mass} density distribution, which can be trivially obtained inverting the barotropic relation, $p = K \rho^{\kappa}$.

For the integration of Eq.~\eqref{eq:equatorial_pot} we use the composite Simpson's rule. It is important to use a very small integration step because the slope is very steep. In this work, we use a step $h = 10^{-6}$. Using the analytic, constant angular momentum case for comparison, we tested that 
a larger value of $h$ gives unacceptable accuracy losses. On the other hand, to integrate the ordinary differential equation \eqref{eq:F} we use a fourth-order Runge-Kutta method. Again, it is also important here to choose a suitable step of integration, especially at the outer end of the disc, because Eq.~\eqref{eq:F} diverges at the equatorial plane (the equipotential surfaces cross the equatorial plane perpendicularly).

@@

The dependence of the location of the maximum of the magnetic pressure with $\beta_{\mathrm{m}_{\mathrm{c}}}$ is shown in the right panel of Fig.~\ref{max-vs-magnetisation}. While this location also decreases with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$, the constant value achieved for values above $\beta_{\mathrm{m}_{\mathrm{c}}}\sim 10^{3}$ does depend on the specific values of $\beta$ and $\gamma$, contrary to what happens with the fluid pressure. It must be noted that the location of the maximum of the magnetic pressure is nearly identical for all models considered when $\log_{10}\beta_{\mathrm{m}_{\mathrm{c}}} \sim 0.5$. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models seem to cross, as it is more clearly shown in the inset of the right panel of Fig.~\ref{max-vs-magnetisation}.

@@

>>>>

Next, we choose all the initial radial values for the integration of Eq.~\eqref{eq:F} to lie between $r_{\mathrm{cusp}}$ and $r_{\mathrm{c}}$ ($\theta = \pi / 2$). Since we are only interested in the equipotential surfaces inside the Roche lobe of the disc, our choice of initial values provides us a mapping of the equipotential surfaces of the torus. Given that we have already obtained both the equipotential surfaces $\theta(r)$ which cross the segment $[r_{\mathrm{cusp}}, r_{\mathrm{c}}]$ at the equatorial plane and the values of the potential in that segment, we can obtain the complete potential distribution for the torus (outside of that segment). Once we have the potential distribution, we can find the {\bf fluid} pressure at the centre of the disc from Eq.~\eqref{eq:enthalpy_eq},
\begin{equation}
p_{\mathrm{c}} = w_{\mathrm{c}}(W_{\mathrm{in}} - W_{\mathrm{c}})\left(\frac{\kappa}{\kappa - 1} + \frac{\lambda
}{\lambda
 - 1}\frac{1}{\beta_{\mathrm{m}_{\mathrm{c}}}}\right)^{-1},
\end{equation}
where $w_{\mathrm{c}}$ is the enthalpy at the centre and
\begin{equation}
\label{eq:beta_eq}
\beta_{\mathrm{m}} = p/p_{\mathrm{m}},
\end{equation}
is the magnetisation parameter ($\beta_{\mathrm{m}_{\mathrm{c}}}$ being the magnetisation parameter at the centre of the disc). Using this definition, we can find the magnetic pressure at the centre,
\begin{equation}
p_{\mathrm{m_{\mathrm{c}}}} = p_{\mathrm{c}}/\beta_{\mathrm{m}_{\mathrm{c}}}\,.
\end{equation}
With both pressures known at the centre, fluid and magnetic, we can now find the constants $K$ and $K_{\mathrm{m}}$ using equations \eqref{eq:eos_fluid} and \eqref{eq:eos_mag}. Therefore, for a given inner radius of the disc $r_{\mathrm{in}}$ we can obtain the potential $W_{\mathrm{in}}$. We now finally have all the ingredients required to find the enthalpy distribution (Eq.~\eqref{eq:enthalpy_eq}), the fluid pressure and magnetic pressure distributions (Eqs.~\eqref{eq:eos_fluid} and \eqref{eq:eos_mag}), and the {\bf rest-mass} density distribution, which can be trivially obtained inverting the barotropic relation, $p = K \rho^{\kappa}$. {\bf We note that the specific internal energy $\varepsilon$ is needed to compute the enthalpy. We obtain
this quantity by combining the ideal fluid and the barotropic equations of state, which yields 
$\varepsilon=\frac{K}{\kappa-1}\rho^{\kappa-1}$.}

For the integration of Eq.~\eqref{eq:equatorial_pot} we use the composite Simpson's rule. It is important to use a very small integration step because the slope is very steep. In this work, we use a step $h = 10^{-6}$. Using the analytic, constant angular momentum case for comparison, we tested that 
a larger value of $h$ gives unacceptable accuracy losses. On the other hand, to integrate the ordinary differential equation \eqref{eq:F} we use a fourth-order Runge-Kutta method. Again, it is also important here to choose a suitable step of integration, especially at the outer end of the disc, because Eq.~\eqref{eq:F} diverges at the equatorial plane (the equipotential surfaces cross the equatorial plane perpendicularly). {\bf We show the proof of this statment in Appendix~\ref{div_partial_W}.}

@@

The dependence of the location of the maximum of the magnetic pressure with $\beta_{\mathrm{m}_{\mathrm{c}}}$ is shown in the right panel of Fig.~\ref{max-vs-magnetisation}. While this location also decreases with decreasing $\beta_{\mathrm{m}_{\mathrm{c}}}$, the constant value achieved for values above $\beta_{\mathrm{m}_{\mathrm{c}}}\sim 10^{3}$ does depend on the specific values of $\beta$ and $\gamma$, contrary to what happens with the fluid pressure. {\bf It must be noted that the location of the maximum of the magnetic pressure is identical for all models considered when $\beta_{\mathrm{m}_{\mathrm{c}}} \equiv 1/(\lambda - 1) = 3$, as we show in Appendix~\ref{app_magmax}. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models cross at $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$, as it is more clearly shown in the inset of the right panel of Fig.~\ref{max-vs-magnetisation}.}

% It must be noted that the location of the maximum of the magnetic pressure is nearly identical for all models considered when $\log_{10}\beta_{\mathrm{m}_{\mathrm{c}}} \sim 0.5$. At this value of $\beta_{\mathrm{m}_{\mathrm{c}}}$ all models seem to cross, as it is more clearly shown in the inset of the right panel of Fig.~\ref{max-vs-magnetisation}.

@@

\begin{appendix}
\section{Divergence of equation~\eqref{eq:F} at $\theta = \pi/2$}\label{div_partial_W}
To prove that Eq.~\eqref{eq:F} diverges at $\theta = \pi/2$ we need to show that
Eq.~\eqref{eq:polar_der_pot} vanishes at the equator, i.e.
\begin{eqnarray}
\partial_{\theta} W = \partial_{\theta} \ln|u_t| - \frac{\Omega \partial_{\theta}l}{1 - \Omega l}\, = 0.
\end{eqnarray}
Then, we have to prove that either the two terms of this equation are equal or they are identically zero at $\theta = \pi/2$. We shall see that the latter is true.
First, we take the derivative of the angular momentum distribution $\partial_{\theta} l$. From Eq.~(7) we can write the angular momentum distribution as $l(r, \theta) = l(r) \sin^{2\gamma} \theta$, so that the derivative reads
\begin{eqnarray}
\partial_{\theta} l = 2\gamma l(r) \sin^{2\gamma - 1} \theta \cos \theta \,.
\end{eqnarray}
Taking $\theta = \pi/2$ leads to $\partial_{\theta} l = 0$, so the second term equals to zero.
To show that the first term is also zero, we write it as
\begin{eqnarray}
\partial_{\theta} \ln|u_t| = \partial_{\theta} \ln \left(\frac{\mathcal{L}}{\mathcal{A}}\right)^{\frac{1}{2}}\,,
\end{eqnarray}
where $\mathcal{L} = g_{t \phi}^2 - g_{tt} g_{\phi\phi}$ and $\mathcal{A} = g_{\phi\phi} + 2 l g_{t\phi} + l^2g_{tt}$, and we have dropped the absolute value, as it is irrelevant to this discussion. Then, the derivative is
\begin{eqnarray}
\partial_{\theta} \ln|u_t| = \frac{1}{2} \frac{\mathcal{A}}{\mathcal{L}}\frac{\mathcal{A}\partial_{\theta}{\mathcal{L}} - \mathcal{L}\partial_{\theta}\mathcal{A}}{\mathcal{A}^2} = \frac{1}{2} \frac{\mathcal{A}\partial_{\theta}{\mathcal{L}} - \mathcal{L}\partial_{\theta}\mathcal{A}}{\mathcal{A} \mathcal{L}}\,.
\end{eqnarray}
We use Boyer-Lindquist coordinates, for which the metric components read
\begin{eqnarray}
g_{tt} = - \left(1 - \frac{2Mr}{\rho^2}\right), \ g_{t\phi} = -\frac{2Mar\sin^2\theta}{\rho^2}, \ g_{\phi\phi} = \left(r^2 + a^2 + \frac{2Ma^2r\sin^2\theta}{\rho^2}\right) \sin^2 \theta\,,
\end{eqnarray}
where $\rho^2 = r^2 + a^2\cos^2 \theta$. Therefore, we obtain $\mathcal{L} = \Delta \sin^2 \theta$, where $\Delta = r^2 - 2Mr + a^2$, and its derivative $\partial_{\theta}\mathcal{L} = 2 \Delta \sin \theta \cos \theta$, which is zero for $\theta = \pi/2$. Next, we take the derivative of $\mathcal{A}$
\begin{eqnarray}
\partial_{\theta} \mathcal{A} = \partial_{\theta} g_{\phi\phi} + 2 (\partial_{\theta} l) g_{t\phi} + 2l\partial_{\theta} g_{t\phi} + 2l(\partial_{\theta} l) g_{tt} + l^2 \partial_{\theta} g_{tt} = \partial_{\theta} g_{\phi\phi} + 2l\partial_{\theta} g_{t\phi} + l^2 \partial_{\theta} g_{tt}\,,
\end{eqnarray}
where we have used the previous result that $\partial_{\theta} l=0$ at $\theta = \pi/2$. By inspecting the metric components, it is easy to see that all terms depending on $\theta$ are functions of $\sin^2 \theta$ or $\cos^2 \theta$. This means their $\theta$ derivatives will have at least a $\cos \theta$ multiplying. Then $\partial_{\theta} \mathcal{A} = 0$ at $\theta = \pi/2$. Therefore, $\partial_{\theta} \ln|u_t|$ is also zero at $\theta = \pi/2$.

\section{Value of the magnetisation parameter for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$}\label{app_magmax}
In this appendix we derive the condition $\beta_{\mathrm{m}_{\mathrm{c}}} = 1/(\lambda - 1)$ for $r_{\mathrm{m}_{\mathrm{max}}} = r_{\mathrm{c}}$.
First, we can use Eq.~\eqref{eq:eos_mag} to write 
\begin{eqnarray}\label{eq_max}
\frac{\partial{p_{\mathrm{m}}(r)}}{\partial r} = \partial_{r} \left(K_{\mathrm{m}} \mathcal{L}^{\lambda -1} w^{\lambda}\right) = 0\,,
\end{eqnarray}
which yields
\begin{eqnarray}
\partial_{r} p_{\mathrm{m}} = K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} [(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w)]=0\,.
\end{eqnarray}
Inside the disc $K_{\mathrm{m}} \mathcal{L}^{\lambda - 2}w^{\lambda} \neq 0$. Therefore, to fulfill the extremum condition~\eqref{eq_max} we need 
\begin{equation}\label{eq:simple_expr}
(\lambda - 1)(\partial_r \mathcal{L}) w + \lambda \mathcal{L} (\partial_r w) = 0\,.
\end{equation}
To evaluate this expression, we need to compute the partial derivatives $\partial_r \mathcal{L}$ and $\partial_r w$. The derivative of $\mathcal{L}$ is straightforward, since $\mathcal{L} = \Delta \sin^2 \theta$,
\begin{equation}
\partial_r \mathcal{L} = 2(r-M)\sin^2 \theta\,.
\end{equation}
Let us now discuss $\partial_r w$. From Eq.~\eqref{eq:enthalpy_eq} we have
\begin{equation}
w = \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{\frac{1}{\lambda-1}}\,.
\end{equation}
Taking its derivative we obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left[\left(-\partial_r W \left(\frac{\lambda -1 }{\lambda}\right) \frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right) + \left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1})^2}\right)\right]\,.
\end{eqnarray}
Since at $r = r_{\mathrm{c}}$ the derivative of the potential is zero, $\partial_r W(r, \pi/2)|_{r = r_{\mathrm{c}}} = 0$, we can simplify the above expression to obtain
\begin{eqnarray}
\partial_r w &=& \frac{w}{\lambda - 1} \left[-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]^{-1} 
\nonumber \\
&\times&
\left(-\Delta W \left(\frac{\lambda -1}{\lambda}\right)\frac{1}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right)\left[\frac{-(\lambda - 1) K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,,
\end{eqnarray}
which can be further simplified to the following form
\begin{equation}
\partial_r w = - w \left[\frac{K_{\mathrm{m}} \mathcal{L}^{\lambda - 2} \partial_r \mathcal{L}}{K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}}\right]\,.
\end{equation}
Next, by inserting this expression in Eq.~\eqref{eq:simple_expr} we obtain
\begin{equation}
(\partial_r \mathcal{L}) w [(\lambda -1)(K+K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}) - \lambda K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}] = 0\,.
\end{equation}
Since $(\partial_r \mathcal{L}) w \neq 0$ inside the disc, the above equation leads to 
\begin{equation}
K (\lambda - 1) = K_{\mathrm{m}}\mathcal{L}^{\lambda - 1}\,.
\end{equation}
Moreover, we can use Eqs.~\eqref{eq:eos_fluid}, \eqref{eq:eos_mag} and \eqref{eq:beta_eq} at the center (remember that $w_{\mathrm{c}} = 1$) to write $K$ as $K = \beta_{\mathrm{m}_{\mathrm{c}}} K_\mathrm{m}\mathcal{L}^{\lambda -1}$. Therefore, this allows us to obtain the following simple expression for the value of the magnetisation parameter at the center of the disc
\begin{equation}
\beta_{\mathrm{m}_{\mathrm{c}}} = \frac{1}{\lambda - 1}\,.
\end{equation}

\sg{We should add a short discussion of the generality of this result.}

\end{appendix}
